Gravitational disturbance

### 2018-04-06 manuscript/manuscript.tex
review up to the normal gravity section

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called gravity vector
and its intensity is called simply gravity 
\citep{hofmann-wellenhof-moritz2005}.

@@

Gravity is the most commonly measured quantity in geophysical surveying.
Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the geoid \citep{li2001}.

@@

@@

%For simplicity, we will henceforth call this 
%resultant gravity corrected from non-gravitational 
%effects due to the vehicle motion and also time variations
%as simply gravity.

In applied geophysics, gravity surveys are usually conducted 
over small areas on the Earth's surface for the purpose of 
characterizing geological structures located within the
crust and upper mantle.

Consequently, geophysicists are only interested in the
particular gravitational component of gravity which 
is produced by these target geological structures.

@@

Traditionally, the Earth's gravity field is approximated 
by the gravity field produced by a geocentric and rigid ellipsoid 
of revolution which has the minor axis $b$ 
coincident with the mean rotating axis of the Earth $Z$, the 
same total mass (including the atmosphere) and also the
same angular velocity of the Earth (Figure \ref{fig:fig1}).
Another characteristic of this model is that its
limiting surface coincides with a particular equipotential 
of its own gravity field.
This model is called as normal Earth and its gravity
field is called normal gravity field
\citep{vanicek1987, hofmann-wellenhof-moritz2005, torge2012}.

@@

The only condition imposed on its internal density
distribution is that it produces a normal gravity field
having a particular equipotential which coincides
with its limiting surface.

For convenience, we denote any density distribution 
satisfying this condition as a normal density distribution.

The normal Earth gives rise to the geodetic coordinate system.
In this coordinate system, the position of a point $P$
is defined by the geometric height $h$, geodetic latitude
$\varphi$ and longitude $\lambda$ (Figure \ref{fig:fig1}).
Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
converted into geocentric Cartesian coordinates $(X, Y, Z)$
(Figure \ref{fig:fig1}).
The plane containing the point $P$, the axis $Z$ and
the origin $O$ of this geocentric Cartesian coordinate system
is called meridian plane (gray plane in Figure \ref{fig:fig1}).

Similarly to the gravity vector and gravity, 
the resultant of the virtual 
gravitational and centrifugal forces exerted by the normal
Earth on a body at rest at a point $P$ is called 
normal gravity vector and its intensity is called 
simply normal gravity.

@@

The gravity anomaly is defined as the difference
between the gravity at the geoid and the normal gravity at the ellipsoid
and is the commonly used quantity in applied geophysics.
Different gravity anomalies can be calculated, depending on the
corrections applied to them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
These corrections are usually called gravity reductions.
For example, the Free-air anomaly is an approximation of the
gravity disturbance whereas the Bouguer anomaly
is an approximation of the terrain corrected gravity disturbance.
The last one is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

@@

By definition, the gravity disturbance $\delta g^{o}_{i}$,
at the point $(x_{i}, y_{i}, z_{i})$, is given by
\begin{equation}
\delta g^{o}_{i} =  \| \mathbf{\delta g}^{o}_{i} \| - \| \boldsymbol{\gamma}_{i} \| \: ,
\label{eq:gravity-disturbance}
\end{equation}
where $\| \mathbf{\delta g}^{o}_{i} \|$ and $\| \boldsymbol{\gamma}_{i} \|$ are,
respectively, the observed gravity and the normal gravity at the
point $(x_{i}, y_{i}, z_{i})$.
Fortunately, the condition $\| \boldsymbol{\gamma}_{i} \| \gg \| \Delta \mathbf{g}^{o}_{i} \|$
is met at all points located above or on the Earth's surface.
By combining this condition and the definition of observed gravity vector
given in equation \ref{eq:gravity-vector}, we can approximate the observed gravity
$\| \mathbf{\delta g}^{o}_{i} \|$ (equation \ref{eq:gravity-disturbance})
by a first order Taylor's expansion as follows:
\begin{equation}
\| \mathbf{\delta g}^{o}_{i} \| \approx \| \boldsymbol{\gamma}_{i} \| + 
\hat{\boldsymbol{\gamma}}_{i}^{\top} \Delta \mathbf{g}^{o}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $\hat{\boldsymbol{\gamma}}_{i}$ is a unit vector with the same 
direction as the normal gravity vector $\boldsymbol{\gamma}_{i}$ at
the point $(x_{i}, y_{i}, z_{i})$.
This approximation is largely used in applied geophysics for representing
total-field anomalies (e.g., \citeauthor{blakely1996}, \citeyear{blakely1996}).
Notice that, local- or regional-gravity studies, the unit
vector $\hat{\boldsymbol{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
coincides with the $z$ axis
of the local Cartesian coordinate system defined
at the beginning of this section. Consequently,
by using the approximation defined in equation \ref{eq:gobs-approx}, 
the gravity disturbance (equation \ref{eq:gravity-disturbance}) can be 
rewritten as follows \begin{equation}
\delta g^{o}_{i} \approx \hat{\mathbf{z}}^{\top} \Delta \mathbf{g}^{o}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mathbf{z}}^{\top} = \left[ \, 0 \; 0 \; 1 \right]$.
According to equation \ref{eq:gravity-disturbance-approx},
the gravity disturbance $\delta g^{o}_{i}$ (equation \ref{eq:gravity-disturbance})
represents the vertical component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a homogeneous gravity source can be represented 
by the following harmonic function
\begin{equation}
d^{o}_{i} = k_{g} \, G \, \rho \, \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $G$ is the Newtonian constant of gravitation
(in $m^{3} / (kg \, s^{2})$),
$k_{g} = 10^{5}$ is a constant factor
transforming from $m/s^{2}$ to milligal (mGal),
and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
$i = 1, ..., N$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
\label{eq:phi}
\end{equation}
with respect to the variable $z$. The 
integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
and $z^{\prime}$ within the volume $v$ of the gravity source.
This equation can be easily generalized for the case
of multiple gravity sources.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.3\textwidth]{figures/geodetic_coordinates.png}
    \caption{Schematic representation of the geodetic coordinate
    system.}
  \label{fig:fig1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.3\textwidth]{figures/gravitational_disturbance.png}
    \caption{Schematic representation of the geodetic coordinate
    system.}
  \label{fig:fig2}
\end{figure}

>>>>

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called 
\textit{gravity vector} and its intensity is called simply 
\textit{gravity} \citep{hofmann-wellenhof-moritz2005}.

@@

Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the geoid \citep{li2001}.

@@

%In applied geophysics, gravity surveys are usually conducted 
%over local or regional areas on the Earth's surface for the purpose of 
%characterizing geological structures located within the
%crust and upper mantle.
%
%Consequently, geophysicists are only interested in the
%particular gravitational component of gravity which 
%is produced by these target geological structures.

@@

@@

Traditionally, the Earth's gravity field is approximated 
by the gravity field produced by a geocentric and rigid ellipsoid 
of revolution which has the minor axis $b$ 
coincident with the mean rotating axis of the Earth $Z$, the 
same total mass (including the atmosphere) and also the
same angular velocity of the Earth (Figure \ref{fig:fig1}).
Another characteristic of this model is that its
limiting surface coincides with a particular equipotential 
of its own gravity field.
This model is called as \textit{normal Earth} 
\citep{vanicek1987, hofmann-wellenhof-moritz2005, torge2012}.

@@

The only condition imposed on its internal density
distribution is that it produces a gravity field
having a particular equipotential which coincides
with its limiting surface.

For convenience, we denote any density distribution 
satisfying this condition as a \textit{normal density distribution}.

Similarly to the gravity vector and gravity, 
the resultant of the virtual 
gravitational and centrifugal forces exerted by the normal
Earth on a body at rest at a point $P$ is called 
\textit{normal gravity vector} and its intensity is called 
simply \textit{normal gravity}.

The normal Earth gives rise to the \textit{geodetic coordinate system}.
In this coordinate system, the position of a point $P$
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude}
$\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
converted into geocentric Cartesian coordinates $(X, Y, Z)$
(Figure \ref{fig:fig1}).

The plane containing the point $P$, the axis $Z$ and
the origin $O$ of this \textit{geocentric Cartesian coordinate system}
is called meridian plane (gray plane in Figure \ref{fig:fig1}).

@@

The gravity anomaly is defined as the difference
between the gravity at the geoid and the normal gravity at the ellipsoid
and is the commonly used quantity in applied geophysics.
Different gravity anomalies can be calculated, depending on the
corrections applied to them \citep{blakely1996, hofmann-wellenhof-moritz2005}.

These corrections are usually called gravity reductions.
For example, the Free-air anomaly is an approximation of the
gravity disturbance whereas the Bouguer anomaly
is an approximation of the terrain corrected gravity disturbance.
The last one is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

@@

By definition, the gravity disturbance $\delta g^{o}_{i}$,
at the point $(x_{i}, y_{i}, z_{i})$, is given by \citep{hofmann-wellenhof-moritz2005}:
\begin{equation}
\delta g^{o}_{i} =  \| \mathbf{\delta g}^{o}_{i} \| - \| \boldsymbol{\gamma}_{i} \| \: ,
\label{eq:gravity-disturbance}
\end{equation}
where $\| \mathbf{\delta g}^{o}_{i} \|$ and $\| \boldsymbol{\gamma}_{i} \|$ are,
respectively, the observed gravity and the normal gravity at the
point $(x_{i}, y_{i}, z_{i})$.
Fortunately, the condition $\| \boldsymbol{\gamma}_{i} \| \gg \| \Delta \mathbf{g}^{o}_{i} \|$
is met at all points located above or on the Earth's surface.
By combining this condition and the definition of observed gravity vector
given in equation \ref{eq:gravity-vector}, we can approximate the observed gravity
$\| \mathbf{\delta g}^{o}_{i} \|$ (equation \ref{eq:gravity-disturbance})
by a first order Taylor's expansion as follows:
\begin{equation}
\| \mathbf{\delta g}^{o}_{i} \| \approx \| \boldsymbol{\gamma}_{i} \| + 
\hat{\boldsymbol{\gamma}}_{i}^{\top} \Delta \mathbf{g}^{o}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $\hat{\boldsymbol{\gamma}}_{i}$ is a unit vector with the same 
direction as the normal gravity vector $\boldsymbol{\gamma}_{i}$ at
the point $(x_{i}, y_{i}, z_{i})$.
This approximation is largely used in applied geophysics for representing
total-field anomalies (e.g., \citeauthor{blakely1996}, \citeyear{blakely1996}).
Notice that, local- or regional-gravity studies, the unit
vector $\hat{\boldsymbol{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
coincides with the $z$ axis
of the local Cartesian coordinate system defined
at the beginning of this section. Consequently,
by using the approximation defined in equation \ref{eq:gobs-approx}, 
the gravity disturbance (equation \ref{eq:gravity-disturbance}) can be 
rewritten as follows \begin{equation}
\delta g^{o}_{i} \approx \hat{\mathbf{z}}^{\top} \Delta \mathbf{g}^{o}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mathbf{z}}^{\top} = \left[ \, 0 \; 0 \; 1 \right]$.
According to equation \ref{eq:gravity-disturbance-approx},
the gravity disturbance $\delta g^{o}_{i}$ (equation \ref{eq:gravity-disturbance})
represents the vertical component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a homogeneous gravity source can be represented 
by the following harmonic function
\begin{equation}
d^{o}_{i} = k_{g} \, G \, \rho \, \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $G$ is the Newtonian constant of gravitation
(in $m^{3} / (kg \, s^{2})$),
$k_{g} = 10^{5}$ is a constant factor
transforming from $m/s^{2}$ to milligal (mGal),
and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
$i = 1, ..., N$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
\label{eq:phi}
\end{equation}
with respect to the variable $z$. The 
integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
and $z^{\prime}$ within the volume $v$ of the gravity source.
This equation can be easily generalized for the case
of multiple gravity sources.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{figures/geodetic_coordinates.png}
    \caption{Schematic representation of the geodetic coordinate
    system defined by an oblate ellipsoid with semi-minor axis $b$, 
    coincident with the mean Earth's rotation axis, and a semi-major
    axis $a$. In this coordinate system, the position of a point is
    determined by the geometric height $h$, the geodetic latitude 
    $\varphi$ and longitude $\lambda$. The Earth's origin is represented 
    by $O$, $P$ represents a point $(h, \varphi, \lambda)$ and 
    $P^{\prime}$ represents the projection of $P$ on the plane $XY$ 
    (Equatorial plane). The plane containing $O$, $P$ and 
    $P^{\prime}$ is represented by the gray triangle in (a) and (b). 
    In (b), the point $Q$ represents the projection of $P$ on the 
    ellipsoid surface.}
  \label{fig:fig1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{figures/gravitational_disturbance.png}
    \caption{Schematic representation of the geodetic coordinate
    system.}
  \label{fig:fig2}
\end{figure}
